---
title: "Dashboard"
date-modified: last-modified
format:
  dashboard:
    theme: [pulse, html/styles.scss, dashboard.scss]
    orientation: rows
    nav-buttons:
      - icon: github
        href: https://github.com/YOUR-USERNAME/YOUR-REPO
        aria-label: GitHub repository for this dashboard
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(here)
library(broom)
library(Epi)
library(gt)
library(plotly)
library(tidyverse)

# Build the analysis data the first time; afterwards just read it
if (!file.exists(here("data", "nh_ldl_bp.rds"))) {
  source(here("R", "prep_nhanes.R"))
}

nh       <- readRDS(here("data", "nh_ldl_bp.rds"))
meta     <- readRDS(here("data", "nh_meta.rds"))
codebook <- readRDS(here("data", "nh_codebook.rds"))

source(here("R", "helpers.R"))
source(here("R", "analyses.R"))

# Convenience objects for the text cards on later pages
b_welch   <- b_results |> filter(str_detect(Method, "Welch"))
d_rr      <- d_results |> filter(Measure == "Relative risk")
tukey_sig <- sum(c_tukey$conf.low > 0 | c_tukey$conf.high < 0)
int_sex   <- interaction_tests |> filter(term == "ldl10:sex")
int_race  <- interaction_tests |> filter(term == "ldl10:race_eth")
stage2    <- e_props |> filter(bp_cat == "Stage 2") |> arrange(pct)
```

# Overview

## Row {height=20%}

```{r}
#| content: valuebox
#| title: "Adults analyzed"
list(icon = "people-fill",     color = "white",   value = fmt(kpi$n, 0))
```

```{r}
#| content: valuebox
#| title: "Median LDL (mg/dL)"
list(icon = "droplet-half",    color = "white",      value = fmt(kpi$median_ldl, 0))
```

```{r}
#| content: valuebox
#| title: "LDL of 130 mg/dL or higher"
list(icon = "arrow-up-circle", color = "white", value = paste0(fmt(kpi$pct_ldl130, 1), "%"))
```

```{r}
#| content: valuebox
#| title: "BP of 130/80 or higher"
list(icon = "heart-pulse",     color = "white",   value = paste0(fmt(kpi$pct_htn, 1), "%"))
```

```{r}
#| content: valuebox
#| title: "LDL-SBP correlation (r)"
list(icon = "graph-up",        color = "white", value = fmt(kpi$r_ldl_sbp, 2))

```

## Row {height=80%}

### Column {width=60%}

```{r}
#| title: "LDL cholesterol vs. systolic blood pressure, by sex (hover for values)"
ggplotly(plot_ldl_sbp_scatter(nh)) |>
  layout(legend = list(orientation = "h", x = 0, y = 1.08))
```

### Column {.tabset width=40%}

::: {.card title="About this dashboard"}
**Question.** Among U.S. adults aged 45-79, is a higher LDL cholesterol level
associated with higher blood pressure, and does that association differ
between men and women or across racial and ethnic groups?

**Audience.** Students, clinicians, and public-health readers who want a
quick, data-grounded look at how two major cardiovascular risk factors
relate to each other.

**How to use it.** Start with the headline numbers above. *Compare means* and *Compare proportions* hold the
formal tests. *Sex & ethnicity* answers the main question. *Data* documents
every variable.

**Data.** NHANES 2017-March 2020 pre-pandemic files, retrieved
`r format(meta$retrieved_on, "%Y-%m-%d")`. Dashboard last rendered
`r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`.
:::

```{r}
#| title: "Systolic BP by LDL category"
#| fig-alt: "Violin and box plots of mean systolic blood pressure for four LDL cholesterol categories."
plot_sbp_by_group(nh, ldl_cat, "LDL category (mg/dL)")
```
```{r}
#| title: "Who is in the sample"
#| fig-alt: "Bar chart of participant counts by race and ethnicity, split by sex."
plot_sample_composition(nh)
```


# Compare means {orientation="columns"}

## Column {width=50%}

### Row {height=55%}

```{r}
#| title: "Analysis B: systolic BP by LDL of 130 mg/dL or higher"
#| fig-alt: "Violin and box plots comparing mean systolic blood pressure between adults with LDL of 130 mg/dL or higher and those below 130."
plot_sbp_by_group(nh, ldl_high, "LDL of 130 mg/dL or higher")
```

### Row {.tabset height=45%}

::: {.card title="Takeaway"}
Difference in mean systolic BP (LDL of 130 or higher minus LDL below 130),
Welch t: **`r fmt_ci(b_welch$estimate, b_welch$conf.low, b_welch$conf.high)` mm Hg**.
The interval `r ci_position(b_welch$conf.low, b_welch$conf.high)` zero.

Group SDs are `r fmt(b_summary$sd[1])` and `r fmt(b_summary$sd[2])` mm Hg with
`r fmt(b_summary$n[1], 0)` and `r fmt(b_summary$n[2], 0)` adults, so Welch's
test (no equal-variance assumption) is the primary result. The bootstrap and
rank-based rows check that the conclusion does not hinge on normality.
:::

```{r}
#| title: "Estimates"
b_results |>
  gt() |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) |>
  fmt_number(columns = p.value, decimals = 3) |>
  sub_missing(missing_text = "-") |>
  cols_label(estimate = "Difference", conf.low = "90% CI low",
             conf.high = "90% CI high", p.value = "p") |>
  tab_source_note("Difference = LDL 130+ minus LDL < 130, in mm Hg. Wilcoxon estimate is the Hodges-Lehmann location shift.")
```

```{r}
#| title: "Group summaries"
b_summary |>
  gt() |>
  fmt_number(columns = c(mean, sd, median), decimals = 1) |>
  cols_label(ldl_high = "LDL 130+", mean = "Mean SBP", sd = "SD", median = "Median SBP")
```

## Column {width=50%}

### Row {height=55%}

```{r}
#| title: "Analysis C: systolic BP across four LDL categories"
#| fig-alt: "Violin and box plots of mean systolic blood pressure across LDL categories below 100, 100 to 129, 130 to 159, and 160 or higher."
plot_sbp_by_group(nh, ldl_cat, "LDL category (mg/dL)")
```

### Row {.tabset height=45%}

::: {.card title="Takeaway"}
One-way ANOVA: F = `r fmt(c_anova$statistic[1], 2)` on
`r c_anova$df[1]` and `r c_anova$df[2]` df, `r fmt_p(c_anova$p.value[1])`.
LDL category accounts for `r fmt(100 * c_r2, 1)`% of the variation in
systolic BP.

`r tukey_sig` of `r nrow(c_tukey)` Tukey HSD pairwise 90% intervals exclude
zero. The Kruskal-Wallis test, which does not assume normality, gives
`r fmt_p(c_kruskal$p.value)`.
:::

```{r}
#| title: "Tukey HSD comparisons"
c_tukey |>
  gt() |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) |>
  fmt_number(columns = adj.p.value, decimals = 3) |>
  cols_label(contrast = "Comparison", estimate = "Difference",
             conf.low = "90% CI low", conf.high = "90% CI high",
             adj.p.value = "Adjusted p")
```

```{r}
#| title: "Group summaries"
c_summary |>
  gt() |>
  fmt_number(columns = c(mean, sd, median), decimals = 1) |>
  cols_label(ldl_cat = "LDL (mg/dL)", mean = "Mean SBP", sd = "SD", median = "Median SBP")
```

# Compare proportions {orientation="columns"}

## Column {width=50%}

### Row {height=55%}

```{r}
#| title: "Analysis D: BP of 130/80 or higher, by LDL of 130 or higher"
#| fig-alt: "Stacked percentage bars showing the share of adults with blood pressure of 130/80 or higher among those with and without LDL of 130 or higher."
plot_stacked_pct(nh, ldl_high, htn_range, pal_yesno,
                 "LDL of 130 mg/dL or higher", "BP 130/80+")
```

### Row {.tabset height=45%}

::: {.card title="Takeaway"}
Share with BP of 130/80 or higher: **`r fmt(100 * d_results$estimate[1], 1)`%**
with LDL of 130 or higher vs. **`r fmt(100 * d_results$estimate[2], 1)`%**
with LDL below 130.

Relative risk: `r fmt_ci(d_rr$estimate, d_rr$conf.low, d_rr$conf.high, 2)`.
The interval `r ci_position(d_rr$conf.low, d_rr$conf.high, null = 1)` 1
(`r fmt_p(d_pvalue)`).
:::

```{r}
#| title: "Risk, RR, OR, RD"
d_results |>
  gt() |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 3) |>
  cols_label(estimate = "Estimate", conf.low = "90% CI low", conf.high = "90% CI high")
```

```{r}
#| title: "2 x 2 counts"
as.data.frame.matrix(d_table) |>
  rownames_to_column("LDL 130+") |>
  gt() |>
  tab_spanner(label = "BP 130/80+", columns = c(Yes, No))
```

## Column {width=50%}

### Row {height=55%}

```{r}
#| title: "Analysis E: ACC/AHA blood pressure category by race/ethnicity"
#| fig-alt: "Stacked percentage bars of blood pressure category (normal or elevated, stage 1, stage 2) for five race and ethnicity groups."
plot_stacked_pct(nh, race_eth, bp_cat, pal_bp, NULL, "BP category") +
  coord_flip() +
  guides(fill = guide_legend(reverse = TRUE))
```

### Row {.tabset height=45%}

::: {.card title="Takeaway"}
Pearson chi-square = `r fmt(e_tidy$statistic, 2)` on `r e_tidy$parameter`
df, `r fmt_p(e_tidy$p.value)`.

Stage 2 BP ranges from `r fmt(stage2$pct[1], 1)`% (`r as.character(stage2$race_eth[1])`)
to `r fmt(last(stage2$pct), 1)`% (`r as.character(last(stage2$race_eth))`). The smallest
cell holds `r e_min_cell` adults (minimum required: 15).
:::

```{r}
#| title: "Counts (row %)"
e_props |>
  mutate(cell = paste0(n, " (", fmt(pct, 0), "%)")) |>
  select(race_eth, bp_cat, cell) |>
  pivot_wider(names_from = bp_cat, values_from = cell) |>
  gt() |>
  cols_label(race_eth = "Race/ethnicity")
```
# Sex & ethnicity

## Row {height=62%}

### Column {width=65%}

```{r}
#| title: "Age-adjusted LDL-SBP slope within each sex and race/ethnicity group"
#| fig-alt: "Forest plot of the change in systolic blood pressure per 10 mg/dL higher LDL, with 90% confidence intervals, for each race and ethnicity group, shown separately for men and women."
plot_strata_forest(strata_slopes)
```

### Column {width=35%}

::: {.card title="Answer to the main question"}
Across all adults, adjusting for age, sex, and race/ethnicity, each 10 mg/dL
higher LDL goes with a
**`r fmt_ci(main_slope$estimate, main_slope$conf.low, main_slope$conf.high, 2)` mm Hg**
difference in mean systolic BP. The interval
`r ci_position(main_slope$conf.low, main_slope$conf.high)` zero.

**Does the slope differ by sex?** Interaction test:
F = `r fmt(int_sex$statistic, 2)`, `r fmt_p(int_sex$p.value)`.

**Does it differ by race/ethnicity?** Interaction test:
F = `r fmt(int_race$statistic, 2)`, `r fmt_p(int_race$p.value)`.

Stratum estimates rest on smaller samples (see table below), so their
intervals are wider; read them together with the interaction tests.
:::

## Row {height=38%}

```{r}
#| title: "Stratum-specific slopes (per 10 mg/dL LDL, age-adjusted)"
strata_slopes |>
  gt() |>
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) |>
  fmt_number(columns = p.value, decimals = 3) |>
  cols_label(sex = "Sex", race_eth = "Race/ethnicity", estimate = "Slope",
             conf.low = "90% CI low", conf.high = "90% CI high", p.value = "p")
```
# Data

## Row {height=60%}

```{r}
#| title: "Codebook"
codebook |>
  gt() |>
  cols_label(Original = "Original NHANES source")
```

## Row {height=40%}

::: {.card title="Sample flow"}
| Step | Participants |
|:-----|-------------:|
| All participants in P_DEMO | `r fmt(meta$n_demo, 0)` |
| Interviewed and examined, age 45-79 | `r fmt(meta$n_eligible, 0)` |
| ...with an LDL result (fasting subsample) | `r fmt(meta$n_with_ldl, 0)` |
| ...and complete data on every variable | `r fmt(meta$n_complete, 0)` |
:::

::: {.card title="About the data"}
**Source.** National Health and Nutrition Examination Survey, 2017-March 2020
pre-pandemic files: P_DEMO (demographics), P_TRIGLY (LDL; measured only in the
morning fasting subsample), and P_BPXO (oscillometric blood pressure).
Retrieved with the nhanesA package on
`r format(meta$retrieved_on, "%Y-%m-%d")`; analysis data prepared
`r format(meta$prepared_on, "%Y-%m-%d")`.

**Caveats.** Survey weights are not used, so results describe this sample,
not the U.S. population. Data are cross-sectional, and medication use
(statins, blood pressure drugs) is not accounted for.

**Accessibility.** Color palettes are safe for common color-vision
deficiencies, every static chart has alt text, and the layout reflows on
phones and tablets.
:::