---
title: "Does LDL Cholesterol Track with Blood Pressure After Age 45?"
subtitle: "Building a NHANES Dashboard to Find Out"
description: "A Quarto dashboard on NHANES 2017-March 2020 data, the hypotheses I tested, and what the data say about LDL, blood pressure, sex, and race/ethnicity."
author: "Liam Wetzel"
date: 2026-09-28
date-modified: last-modified
date-format: iso
categories: [NHANES, Quarto, R, cardiovascular health]
---

```{r}
#| label: setup
#| include: false

library(here)
library(broom)
library(Epi)
library(tidyverse)

nh   <- readRDS(here("data", "nh_ldl_bp.rds"))
meta <- readRDS(here("data", "nh_meta.rds"))

source(here("R", "helpers.R"))
source(here("R", "analyses.R"))

# Shortcuts to single results used in the text
b_welch   <- b_results |> filter(str_detect(Method, "Welch"))
d_rr      <- d_results |> filter(Measure == "Relative risk")
tukey_sig <- sum(c_tukey$conf.low > 0 | c_tukey$conf.high < 0)
int_sex   <- interaction_tests |> filter(term == "ldl10:sex")
int_race  <- interaction_tests |> filter(term == "ldl10:race_eth")
stage2    <- e_props |> filter(bp_cat == "Stage 2") |> arrange(pct)

# Extra results that appear only in this post
age_cor <- nh |>
  summarise(r_age_ldl = cor(age, ldl), r_age_sbp = cor(age, sbp))

sex_slopes <- nh |>
  group_by(sex) |>
  group_modify(~ tidy(lm(sbp ~ ldl10 + age + race_eth, data = .x),
                      conf.int = TRUE, conf.level = conf_level)) |>
  ungroup() |>
  filter(term == "ldl10")

male   <- sex_slopes |> filter(sex == "Male")
female <- sex_slopes |> filter(sex == "Female")
```


## Introduction

High LDL cholesterol and high blood pressure are two of the most important
modifiable risk factors for heart attack and stroke, and they often show up in
the same people. I wanted to know whether, in a large sample of U.S. adults,
people with higher LDL also tend to have higher blood pressure, or whether the
two risk factors travel more independently than we might assume.

I focused on adults aged 45 to 79 for two reasons. Both LDL and blood pressure
typically climb through midlife, and this is the age range where women's
cardiovascular risk profile shifts around menopause, which makes a comparison
of men and women especially interesting. I also compared racial and ethnic
groups, since the burden of hypertension in the United States is not evenly
shared.

## The Data

I used the National Health and Nutrition Examination Survey (NHANES)
2017-March 2020 pre-pandemic files, pulled into R with the **nhanesA**
package on `r format(meta$retrieved_on, "%B %d, %Y")`. Three tables were
merged on the participant ID (SEQN):

- **P_DEMO** for age, sex, and race/ethnicity
- **P_TRIGLY** for LDL cholesterol (Friedewald equation, mg/dL)
- **P_BPXO** for oscillometric blood pressure, averaged over two or three readings

LDL is measured only in the morning fasting subsample, which is the main
reason the final sample is smaller than the full survey:

| Step | Participants |
|:-----|-------------:|
| All participants in P_DEMO | `r fmt(meta$n_demo, 0)` |
| Interviewed and examined, age 45-79 | `r fmt(meta$n_eligible, 0)` |
| ...with an LDL result (fasting subsample) | `r fmt(meta$n_with_ldl, 0)` |
| ...and complete data on every variable | `r fmt(meta$n_complete, 0)` |

Participants aged 80 were excluded because NHANES codes everyone aged 80 and older
as "80". One participant had an LDL of `r min(nh$ldl)` mg/dL; I kept this
value, since very low LDL is plausible with intensive lipid-lowering therapy.
Survey weights were not used, so the results describe this sample rather than
the U.S. population as a whole.

```{r}
#| label: show-prep-code
#| echo: true
#| eval: false
#| code-fold: true
#| code-summary: "Show the data ingest and cleaning code"
#| file: ../../R/prep_nhanes.R
```

## Hypotheses

Before looking at the results, I wrote down what I expected:

1. Adults with LDL of 130 mg/dL or higher have higher mean systolic blood
   pressure than adults with lower LDL.
2. Mean systolic blood pressure rises steadily across LDL categories
   (below 100, 100-129, 130-159, and 160 or higher).
3. Blood pressure of 130/80 or higher is more common among adults with LDL of
   130 or higher.
4. The mix of blood pressure categories differs across racial and ethnic groups.
5. The LDL-blood pressure relationship is stronger in women than in men, and
   differs across racial and ethnic groups.


## Results

All intervals are 90% confidence intervals.

**Higher LDL did not mean higher average systolic blood pressure, at first
glance.** Comparing adults with LDL of 130 or higher to everyone else, the
difference in mean systolic blood pressure was
`r fmt_ci(b_welch$estimate, b_welch$conf.low, b_welch$conf.high)` mm Hg
(Welch's t test). The interval `r ci_position(b_welch$conf.low, b_welch$conf.high)`
zero, and pooled t, bootstrap, and Wilcoxon rank sum estimates all told the
same story, so this result does not hinge on a normality assumption. Across
four LDL categories, a one-way ANOVA gave F = `r fmt(c_anova$statistic[1], 2)`
(`r fmt_p(c_anova$p.value[1])`), with `r tukey_sig` of
`r nrow(c_tukey)` Tukey comparisons excluding zero, but LDL category explained
only `r fmt(100 * c_r2, 1)`% of the variation in systolic blood pressure, so
even where some groups differ, LDL category on its own tells us very little
about a person's blood pressure. Hypothesis 1 was not supported by this
unadjusted comparison.

**Hypertensive-range blood pressure was modestly more common with high LDL.**
Blood pressure of 130/80 or higher was found in
`r fmt(100 * d_results$estimate[1], 1)`% of adults with LDL of 130 or higher
and `r fmt(100 * d_results$estimate[2], 1)`% of those with lower LDL, a
relative risk of `r fmt_ci(d_rr$estimate, d_rr$conf.low, d_rr$conf.high, 2)`.
This supports hypothesis 3. Why would this analysis find an association when
the comparison of means did not? The 130/80 threshold also counts diastolic
pressure, and it asks about crossing a cutoff rather than about the average,
so a shift near the cutoff can show up here without moving the mean much.

**Blood pressure categories differed across racial and ethnic groups.** The
chi-square test of independence gave
`r fmt(e_tidy$statistic, 1)` on `r e_tidy$parameter` df
(`r fmt_p(e_tidy$p.value)`), supporting hypothesis 4. Stage 2 blood pressure
ranged from `r fmt(stage2$pct[1], 1)`% of
`r as.character(stage2$race_eth[1])` adults to
`r fmt(last(stage2$pct), 1)`% of `r as.character(last(stage2$race_eth))`
adults.

```{r}
#| label: fig-bp-by-race
#| fig-cap: "Blood pressure category by race/ethnicity"
#| fig-alt: "Stacked percentage bars of blood pressure category for five race and ethnicity groups."
plot_stacked_pct(nh, race_eth, bp_cat, pal_bp, NULL, "BP category") +
  coord_flip() +
  guides(fill = guide_legend(reverse = TRUE))
```

**Age was hiding the LDL-blood pressure association.** Older participants
tended to have lower LDL (r = `r fmt(age_cor$r_age_ldl, 2)`) but higher
systolic blood pressure (r = `r fmt(age_cor$r_age_sbp, 2)`). Because age
pushes the two in opposite directions, it masks their relationship when age
is ignored. After adjusting for age, sex, and race/ethnicity, each 10 mg/dL
higher LDL went with a
`r fmt_ci(main_slope$estimate, main_slope$conf.low, main_slope$conf.high, 2)`
mm Hg higher mean systolic blood pressure. That is small per unit, but across
a 100 mg/dL range of LDL it adds up to about
`r fmt(10 * main_slope$estimate, 1)` mm Hg.

**The relationship was steeper in men, the opposite of what I expected.** The
age-adjusted slope was `r fmt_ci(male$estimate, male$conf.low, male$conf.high, 2)`
mm Hg per 10 mg/dL in men and
`r fmt_ci(female$estimate, female$conf.low, female$conf.high, 2)` in women.
The test that directly compares the two slopes gave
`r fmt_p(int_sex$p.value)`, which is modest evidence of a difference at my
90% standard. There was no evidence that the slope differed across racial and
ethnic groups (`r fmt_p(int_race$p.value)`). So hypothesis 5 was not
supported: the sex difference ran in the other direction, and the ethnic
differences did not appear.

```{r}
#| label: fig-forest
#| fig-cap: "Age-adjusted change in systolic BP per 10 mg/dL higher LDL, by sex and race/ethnicity"
#| fig-alt: "Forest plot of LDL-systolic blood pressure slopes with 90 percent confidence intervals for each race and ethnicity group, shown separately for men and women."
#| fig-height: 4.5
plot_strata_forest(strata_slopes)
```

## Conclusions

In adults aged 45 to 79, higher LDL goes with slightly higher blood pressure,
but the association only becomes clear once age is taken into account, and it
appears stronger in men than in women. The effect is modest for any one person,
yet at the population level a few mm Hg of systolic pressure is meaningful.
I can only guess at why the sex difference runs the way it does: statin use
may differ between men and women, and changes around menopause may shift LDL
and blood pressure somewhat independently. Both are ideas to test, not
conclusions.

## Limitations

- **Cross-sectional data.** Each person was measured once, so these are
  associations, not effects of LDL on blood pressure.
- **Medication use.** Many adults in this age range take statins or blood
  pressure drugs. If people with high blood pressure are more likely to be on a
  statin, treatment could weaken or even reverse the LDL-blood pressure
  association seen here. The NHANES blood pressure and cholesterol
  questionnaire (P_BPQ) would allow adjusting for this.
- **Fasting subsample, no weights.** LDL comes from a morning subsample, and I
  did not apply survey weights, so the results are not nationally
  representative.
- **Smaller groups.** Estimates for NH Asian and Other/Multiracial adults rest
  on fewer participants and have wider intervals.

## Future Analyses

The natural next step is to add medication use from P_BPQ and refit the
models, and to look at diastolic blood pressure alongside systolic. I would
also like to try the Martin-Hopkins LDL estimate, which NHANES also provides
and which is more accurate than the Friedewald equation when triglycerides are
high.

## About the Dashboard

The dashboard is a Quarto file using `format: dashboard`, published with the
rest of this site on GitHub Pages. Data cleaning, statistics, and plotting
each live in their own R script, and this post reads from the same analysis
script, so the numbers always match. The Overview page leads with five
headline numbers and an interactive scatter plot. Each analysis page pairs a
chart with tabs for the takeaway, the estimates, and the group summaries, and
a Data page documents every variable along with retrieval and render dates.
Colors are safe for color-blind readers (from the Okabe-Ito palette), and every static chart has alt text.